---
title: "March Madness Descriptives"
subtitle: "Regular Season"
author: "Ekarin Eric Pongpipat, M.A."
date: "Date"
output:
  html_document:
    highlight: textmate
    theme: lumen
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

## R Packages
```{r}
packages <- c("tidyverse", "kableExtra", "broom", "broomExtra", "gganimate")
xfun::pkg_attach(packages, message = F)
```

## Functions
```{r}
nice_tables <- function(table) {
  table %>%
    kable() %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width = "100%", height = "200px")
}
```

Let's first look at the regular season dataset.

## Regular Season
```{r}
df_regular_season <- read_csv("data/RegularSeasonDetailedResults.csv")
glimpse(df_regular_season)
```

We will also need to load the Teams.csv dataset to append actual team names rather than IDs.

```{r}
df_teams <- read_csv("data/Teams.csv")
glimpse(df_teams)

# select only the TeamID and TeamName vars
df_teams <- df_teams %>% select(TeamID, TeamName) 

# create df for win and loss TeamID and TeamNames to append to regular season df
df_wteams <- df_teams                         
colnames(df_wteams) <- c("WTeamID", "WTeamName")
df_lteams <- df_teams                        
colnames(df_lteams) <- c("LTeamID", "LTeamName")

# add actual names using inner_join
# remove TeamID
df_regular_season <- df_regular_season %>%
  inner_join(., df_wteams, by = "WTeamID") %>%
  inner_join(., df_lteams, by = "LTeamID") %>%
  select(-contains("TeamID"))

glimpse(df_regular_season)
```

Let's break apart the tables into wins and losses so that we can manipulate the same variables at the same time.

```{r}
df_regular_season_win <- df_regular_season %>%
  select(Season, DayNum, contains("W"))
colnames(df_regular_season_win) <- gsub("W", "", colnames(df_regular_season_win))

df_regular_season_loss <- df_regular_season %>%
  select(Season, DayNum, contains("L"))
colnames(df_regular_season_loss) <- gsub("L", "", colnames(df_regular_season_loss))

# data manipulation ----
# create new variables that are also easy to read
data_manipulation_reg_season <- function(data) {
  data %>%
  select(season = Season,
         n_day = DayNum,
         team_name = TeamName,
         score = Score,
         points_made = FGM,
         points_attempted = FGA,
         points_3_made = FGM3,
         points_3_attempted = FGA3,
         points_free_throws_made = FTM,
         points_free_throws_attempted = FTA,
         rebound_offensive = OR,
         rebound_defensive = DR,
         assist = Ast,
         turnovers = TO,
         steals = Stl,
         blocks = Blk,
         personal_fouls = PF) %>%
  mutate(index = 1:nrow(data),
         points_2_made = points_made - points_3_made,
         points_2_attempted = points_attempted - points_3_attempted,
         points_missed = points_attempted - points_made,
         points_2_missed = points_2_attempted - points_2_made,
         points_3_missed = points_3_attempted - points_3_made,
         points_free_throws_missed = points_free_throws_attempted - points_free_throws_made,
         points_made_percentage = points_made / points_attempted * 100,
         points_2_made_percentage = points_2_made / points_2_attempted * 100,
         points_3_made_percentage = points_3_made / points_3_attempted * 100,
         points_free_throws_made_percentage = points_free_throws_made / points_free_throws_attempted * 100)
}

df_regular_season_win <- data_manipulation_reg_season(df_regular_season_win) %>% mutate(outcome = "win")
glimpse(df_regular_season_win)

df_regular_season_loss <- data_manipulation_reg_season(df_regular_season_loss) %>% mutate(outcome = "loss")
glimpse(df_regular_season_loss)

df_regular_season_all <- bind_rows(df_regular_season_win, df_regular_season_loss) %>% mutate(outcome = as_factor(outcome))
glimpse(df_regular_season_all)
```

Now let's compare some winning and losing stats.

## Across All Seasons

### Most Wins

```{r}
df_regular_season_all %>%
  group_by(team_name, outcome) %>%
  count() %>%
  ungroup() %>%
  filter(outcome == "win") %>%
  top_n(10) %>%
  arrange(desc(n)) %>%
  mutate(team_name = factor(team_name, levels = team_name)) %>%
  ggplot(., aes(x = team_name, y = n, fill = n)) + 
    geom_col() +
    scale_fill_distiller(palette = "Blues", direction = -1) +
    theme_minimal() + 
    theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
    labs(title = "Most Wins During Regular Season",
         x = "\nTeam",
         y = "Frequency\n",
         fill = "Frequency\n")
```

### Metrics
```{r}
reg_season_metrics <- df_regular_season_all %>%
  select(-c(season, n_day, team_name, index, outcome)) %>%
  colnames()

for (metric in reg_season_metrics) {
  figure <- df_regular_season_all %>%
    ggplot(., aes(x = eval(as.name((metric))), fill = outcome)) + 
      geom_density(alpha = 0.5, color = 0) +
      theme_minimal() + 
      labs(x = paste0("\n",metric),
           y = "Density\n")
  print(figure)
}
```

It looks like the winning team takes slightly less shots, misses less, has more offensive rebounds, more assists, less turnovers, more steals, more blocks, and less personal fouls.

## By Season

### Most Wins
```{r, fig.width = 12, fig.height = 12}
df_most_wins <- df_regular_season_all %>%
  group_by(season, team_name, outcome) %>%
  count() %>%
  ungroup() %>%
  filter(outcome == "win") %>%
  group_by(season) %>%
  top_n(10) %>%
  ungroup() %>%
  arrange(season, desc(n)) %>%
  mutate(order = row_number())

ggplot(df_most_wins, aes(x = order, y = n, fill = n)) + 
    geom_col() +
    scale_fill_distiller(palette = "Blues", direction = -1) +
    theme_minimal() + 
    theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
    labs(title = "Most Wins During Regular Season",
         x = "\nTeam",
         y = "Frequency\n",
         fill = "Frequency\n") +
  facet_wrap(~ season, scales = "free") +
  scale_x_continuous(breaks = df_most_wins$order,
                     labels = df_most_wins$team_name,
                     expand = c(0,0))
```

### Metrics

```{r}
for (metric in reg_season_metrics) {
  figure <- df_regular_season_all %>%
    ggplot(., aes(x = eval(as.name((metric))), fill = outcome)) + 
      geom_density(alpha = 0.5, color = 0) +
      theme_minimal() + 
      labs(x = paste0("\n",metric),
           y = "Density\n") + 
      facet_wrap(~ season)
  print(figure)
}
```

These metrics look pretty consistent across the seasons.