---
title: "March Madness 2019"
author: "Ekarin Eric Pongpipat, M.A."
date: "Date"
output:
  html_document:
    highlight: textmate
    theme: lumen
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

## R Packages
```{r}
packages <- c("tidyverse", "kableExtra", "broom", "broomExtra", "gganimate")
xfun::pkg_attach(packages, message = F)
```

## Functions
```{r}
nice_tables <- function(table) {
  table %>%
    kable() %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width = "100%", height = "200px")
}
```

## Regular Season

* FGM - field goals made
* FGA - field goals attempted
* FGM3 - three pointers made
* FGA3 - three pointers attempted
* FTM - free throws made
* FTA - free throws attempted
* OR - offensive rebounds
* DR - defensive rebounds
* Ast - assists
* TO - turnovers committed
* Stl - steals
* Blk - blocks
* PF - personal fouls committed

```{r}
df_regular_season <- read_csv("data/RegularSeasonDetailedResults.csv")

df_regular_season <- df_regular_season %>%
  rowwise() %>%
  mutate(WFGM2 = WFGM-WFGM3,
         WFGA2 = WFGA-WFGA3,
         LFGM2 = LFGM-LFGM3,
         LFGA2 = LFGM-LFGA3,
         diff_score = WScore-LScore,
         team_id_pair = paste0(sort(c(WTeamID, LTeamID)), collapse = "_")) %>%
  select(-DayNum, -WTeamID, -WScore, -LTeamID, -LScore, -WLoc, -NumOT, -WFGM, -WFGA, -LFGM, -LFGA) %>%
  ungroup()

df_regular_season[2:28] <- apply(df_regular_season[2:28], 2, scale)
```

### Summary
```{r}
summary(df_regular_season) %>%
  nice_tables() %>%
  scroll_box(width = "100%")
```

### Correlation Matrix
```{r}
df_reg_season_nested <- df_regular_season %>%
  select(-team_id_pair) %>%
  group_by(Season) %>%
  nest() %>%
  mutate(cor_mat = map(data, cor),
         cor_mat_long = map(cor_mat, reshape::melt)) %>%
  select(-data, -cor_mat) %>%
  unnest()

ggplot(df_reg_season_nested, aes(X1, X2, fill = value)) +
  geom_raster() +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
  labs(title = 'Year: {round(frame_time,0)}',
       x = "Metric",
       y = "Metric",
       fill = "Correlation") +
  scale_fill_distiller(palette = "RdBu", limits = c(-1, 1)) +
  transition_time(Season)
```

### First-Level Analysis
```{r}
df_level1_tidy <- df_regular_season %>%
  select(-team_id_pair) %>%
  group_by(Season) %>%
  do(level1 = lm(diff_score ~ ., data = .)) %>%
  tidy(level1) %>%
  mutate(p.value.fdr = p.adjust(p.value, "fdr")) %>%
  tidy_lm_add_logical_significance(tidy_df = ., p_values = "p.value.fdr", sig_column_name = "sig.fdr.05")

df_level1_tidy %>% 
  nice_tables()
```

#### Significant Only
```{r}
df_level1_tidy %>% 
  filter(sig.fdr.05 == TRUE) %>% 
  nice_tables()
```

### Second-Level Analysis
```{r}
colnames(df_level1_tidy)[2] <- "term_level1"
df_level2_tidy <- df_level1_tidy %>%
  group_by(term_level1) %>%
  do(level2 = lm(estimate ~ Season, .)) %>%
  tidy(level2) %>%
  mutate(p.value.fdr = p.adjust(p.value, "fdr")) %>%
  tidy_lm_add_logical_significance(tidy_df = ., p_values = "p.value.fdr", sig_column_name = "sig.fdr.05") %>%
  tidy_lm_add_r_squared(., n = nrow(df_level1_tidy))

df_level2_tidy %>% nice_tables()
```

#### Significant Only
```{r}
df_level2_tidy %>% 
  filter(sig.fdr.05 == TRUE) %>% 
  nice_tables()
```

## NCAA Tourney
```{r}
df_ncaa_tourney <- read_csv("data/NCAATourneyDetailedResults.csv")
df_ncaa_tourney <- df_ncaa_tourney %>%
  rowwise() %>%
  mutate(WFGM2 = WFGM-WFGM3,
         WFGA2 = WFGA-WFGA3,
         LFGM2 = LFGM-LFGM3,
         LFGA2 = LFGM-LFGA3,
         diff_score = WScore-LScore,
         team_id_pair = paste0(sort(c(WTeamID, LTeamID)), collapse = "_")) %>%
  select(-DayNum, -WTeamID, -WScore, -LTeamID, -LScore, -WLoc, -NumOT, -WFGM, -WFGA, -LFGM, -LFGA) %>%
  ungroup()

df_ncaa_tourney[2:28] <- apply(df_ncaa_tourney[2:28], 2, scale)
```

### Summary
```{r}
summary(df_ncaa_tourney) %>%
  nice_tables() %>%
  scroll_box(width = "100%")
```

### Correlation Matrix
```{r}
df_ncaa_tourney  %>%
  select(-team_id_pair) %>%
  cor() %>%
  reshape2::melt(., id = "Season") %>%
  ggplot(., aes(Var1, Var2, fill = value)) +
  geom_raster() +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
  labs(x = "Metric",
       y = "Metric",
       fill = "Correlation") +
  scale_fill_distiller(palette = "RdBu", limits = c(-1, 1))
```

### First-Level Analysis
```{r}
df_level1_tidy <- df_ncaa_tourney %>%
  select(-team_id_pair) %>%
  group_by(Season) %>%
  do(level1 = lm(diff_score ~ ., data = .)) %>%
  tidy(level1) %>%
  mutate(p.value.fdr = p.adjust(p.value, "fdr")) %>%
  tidy_lm_add_logical_significance(tidy_df = ., p_values = "p.value.fdr", sig_column_name = "sig.fdr.05")

df_level1_tidy %>% 
  nice_tables()
```

#### Significant Only
```{r}
df_level1_tidy %>% 
  filter(sig.fdr.05 == TRUE) %>% 
  nice_tables()
```

### Second-Level Analysis
```{r}
colnames(df_level1_tidy)[2] <- "term_level1"
df_level2_tidy <- df_level1_tidy %>%
  group_by(term_level1) %>%
  do(level2 = lm(estimate ~ Season, .)) %>%
  tidy(level2) %>%
  mutate(p.value.fdr = p.adjust(p.value, "fdr")) %>%
  tidy_lm_add_logical_significance(tidy_df = ., p_values = "p.value.fdr", sig_column_name = "sig.fdr.05") %>%
  tidy_lm_add_r_squared(., n = nrow(df_level1_tidy))

df_level2_tidy %>% nice_tables()
```

#### Significant Only
```{r}
df_level2_tidy %>% 
  filter(sig.fdr.05 == TRUE) %>% 
  nice_tables()
```

## Combined
```{r}
df_regular_season <- df_regular_season %>%
  mutate(type = -0.5)
df_ncaa_tourney <- df_ncaa_tourney %>%
  mutate(type = 0.5)

df_combined <- rbind(df_regular_season, df_ncaa_tourney)
```

### Summary
```{r}
summary(df_combined) %>%
  nice_tables() %>%
  scroll_box(width = "100%")
```

### Correlation Matrix
```{r}
df_combined  %>%
  select(-team_id_pair, -type) %>%
  cor() %>%
  reshape2::melt(., id = "Season") %>%
  ggplot(., aes(Var1, Var2, fill = value)) +
  geom_raster() +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
  labs(x = "Metric",
       y = "Metric",
       fill = "Correlation") +
  scale_fill_distiller(palette = "RdBu", limits = c(-1, 1))
```

### First-Level Analysis
```{r}
df_level1_tidy <- df_combined %>%
  select(-team_id_pair) %>%
  group_by(Season) %>%
  do(level1 = lm(diff_score ~ ., data = .)) %>%
  tidy(level1) %>%
  mutate(p.value.fdr = p.adjust(p.value, "fdr")) %>%
  tidy_lm_add_logical_significance(tidy_df = ., p_values = "p.value.fdr", sig_column_name = "sig.fdr.05")

df_level1_tidy %>% 
  nice_tables()
```

#### Significant Only
```{r}
df_level1_tidy %>% 
  filter(sig.fdr.05 == TRUE) %>% 
  nice_tables()
```

### Second-Level Analysis
```{r}
colnames(df_level1_tidy)[2] <- "term_level1"
df_level2_tidy <- df_level1_tidy %>%
  group_by(term_level1) %>%
  do(level2 = lm(estimate ~ Season, .)) %>%
  tidy(level2) %>%
  mutate(p.value.fdr = p.adjust(p.value, "fdr")) %>%
  tidy_lm_add_logical_significance(tidy_df = ., p_values = "p.value.fdr", sig_column_name = "sig.fdr.05") %>%
  tidy_lm_add_r_squared(., n = nrow(df_level1_tidy))

df_level2_tidy %>% nice_tables()
```

#### Significant Only
```{r}
df_level2_tidy %>% 
  filter(sig.fdr.05 == TRUE) %>% 
  nice_tables()
```

```{r}

```