---
title: "March Madness Analysis"
subtitle: "Tournament by Regular Season - Round 5"
author: "Ekarin Eric Pongpipat, M.A."
date: "Date"
output:
  html_document:
    highlight: textmate
    theme: lumen
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

## R Packages
```{r}
packages <- c("tidyverse", "kableExtra", "broom", "broomExtra", "gganimate", "caret", "glmnet")
xfun::pkg_attach(packages, message = F)
```

## Functions
```{r}
# create nice html tables ----
nice_tables <- function(table) {
  table %>%
    kable() %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width = "100%", height = "200px")
}

# metric data manipulation ----
metric_data_manipulation <- function(data) {
  data %>%
  select(season = Season,
         n_day = DayNum,
         team_name = TeamName,
         score = Score,
         points_made = FGM,
         points_attempted = FGA,
         points_3_made = FGM3,
         points_3_attempted = FGA3,
         points_free_throws_made = FTM,
         points_free_throws_attempted = FTA,
         rebound_offensive = OR,
         rebound_defensive = DR,
         assist = Ast,
         turnovers = TO,
         steals = Stl,
         blocks = Blk,
         personal_fouls = PF) %>%
  mutate(index = 1:nrow(data),
         points_2_made = points_made - points_3_made,
         points_2_attempted = points_attempted - points_3_attempted,
         points_missed = points_attempted - points_made,
         points_2_missed = points_2_attempted - points_2_made,
         points_3_missed = points_3_attempted - points_3_made,
         points_free_throws_missed = points_free_throws_attempted - points_free_throws_made,
         points_made_percentage = points_made / points_attempted * 100,
         points_2_made_percentage = points_2_made / points_2_attempted * 100,
         points_3_made_percentage = points_3_made / points_3_attempted * 100,
         points_free_throws_made_percentage = points_free_throws_made / points_free_throws_attempted * 100)
}
```

## Datasets

### Regular Season (Before 2018)
```{r}
df_regular_season <- read_csv("data/RegularSeasonDetailedResults.csv")

# add TeaamID and TeamName dataset
df_teams <- read_csv("data/Teams.csv")

# select only the TeamID and TeamName vars
df_teams <- df_teams %>% 
  select(TeamID, TeamName) 

# create df for win and loss TeamID and TeamNames to append to regular season df
df_wteams <- df_teams                         
colnames(df_wteams) <- c("WTeamID", "WTeamName")
df_lteams <- df_teams                        
colnames(df_lteams) <- c("LTeamID", "LTeamName")

# add actual names using inner_join
# remove TeamID
df_regular_season <- df_regular_season %>%
  inner_join(., df_wteams, by = "WTeamID") %>%
  inner_join(., df_lteams, by = "LTeamID") %>%
  select(-contains("TeamID"))

glimpse(df_regular_season)
```

Let's break apart the tables into wins and losses so that we can manipulate the same variables at the same time.

```{r}
# separate wins and losses
df_regular_season_win <- df_regular_season %>%
  select(Season, DayNum, contains("W"))
colnames(df_regular_season_win) <- gsub("W", "", colnames(df_regular_season_win))

df_regular_season_loss <- df_regular_season %>%
  select(Season, DayNum, contains("L"))
colnames(df_regular_season_loss) <- gsub("L", "", colnames(df_regular_season_loss))

# include outcome
df_regular_season_win <- metric_data_manipulation(df_regular_season_win) %>% 
  mutate(outcome = "win")

df_regular_season_loss <- metric_data_manipulation(df_regular_season_loss) %>% 
  mutate(outcome = "loss")

df_win_num <- df_regular_season_win %>%
  select(-c(season, n_day, team_name, outcome, index))
    
df_loss_num <- df_regular_season_loss %>%
  select(-c(season, n_day, team_name, outcome, index))

df_reg_season_diff <- df_win_num-df_loss_num

df_reg_season_diff <- df_reg_season_diff %>%
  mutate(season = df_regular_season_win$season,
         n_day = df_regular_season_win$n_day,
         team_name = df_regular_season_win$team_name,
         index = df_regular_season_win$index)
glimpse(df_reg_season_diff)

# combine together using bind_rows 
df_regular_season_all <- bind_rows(df_regular_season_win, df_regular_season_loss) %>% 
  mutate(outcome = as_factor(outcome))

glimpse(df_regular_season_all)
```

Let's obtain overall season stats for each team. Let's obtain the total and mean.

```{r}
df_regular_season_all_total <- df_regular_season_all %>%
  select(-n_day, -index) %>%
  rowwise() %>%
  mutate(n_win = ifelse(outcome == "win", 1, 0)) %>%
  ungroup() %>%
  select(-outcome) %>%
  group_by(season, team_name) %>%
  summarise_all(sum) %>%
  ungroup()
glimpse(df_regular_season_all_total)

df_regular_season_all_mean <- df_regular_season_all %>%
  select(-n_day, -index, -outcome) %>%
  group_by(season, team_name) %>%
  summarise_all(mean) %>%
  ungroup()
glimpse(df_regular_season_all_mean)

df_reg_season_diff_total <- df_reg_season_diff %>%
  select(-c(index, n_day)) %>%
  group_by(season, team_name) %>%
  summarise_all(sum)
glimpse(df_reg_season_diff_total)

df_reg_season_diff_mean <- df_reg_season_diff %>%
  select(-c(index, n_day)) %>%
  group_by(season, team_name) %>%
  summarise_all(mean)
glimpse(df_reg_season_diff_mean)

colnames(df_regular_season_all_total)[3:ncol(df_regular_season_all_total)] <- paste0(colnames(df_regular_season_all_total)[3:ncol(df_regular_season_all_total)], "_total")

colnames(df_regular_season_all_mean)[3:ncol(df_regular_season_all_mean)] <- paste0(colnames(df_regular_season_all_mean)[3:ncol(df_regular_season_all_mean)], "_mean")

colnames(df_reg_season_diff_total)[3:ncol(df_reg_season_diff_total)] <- paste0(colnames(df_reg_season_diff_total)[3:ncol(df_reg_season_diff_total)], "_diff_total")

colnames(df_reg_season_diff_mean)[3:ncol(df_reg_season_diff_mean)] <- paste0(colnames(df_reg_season_diff_mean)[3:ncol(df_reg_season_diff_mean)], "_diff_mean")
```

### Tourney (Before 2018)
```{r}
df_tournament <- read_csv("data/NCAATourneyDetailedResults.csv") %>%
  inner_join(., df_wteams, by = "WTeamID") %>%
  inner_join(., df_lteams, by = "LTeamID") %>%
  select(-contains("TeamID"))
glimpse(df_tournament)

df_tournament_win <- df_tournament %>%
  select(Season, DayNum, contains("W"))
colnames(df_tournament_win) <- gsub("W", "", colnames(df_tournament_win))

df_tournament_loss <- df_tournament %>%
  select(Season, DayNum, contains("L"))
colnames(df_tournament_loss) <- gsub("L", "", colnames(df_tournament_loss))

df_tournament_win <- metric_data_manipulation(df_tournament_win) %>% 
  mutate(outcome = "win")
glimpse(df_tournament_win)

df_tournament_loss <- metric_data_manipulation(df_tournament_loss) %>% 
  mutate(outcome = "loss")
glimpse(df_tournament_loss)

df_tournament_all <- bind_rows(df_tournament_win, df_tournament_loss) %>% 
  mutate(outcome = as_factor(outcome))
glimpse(df_tournament_all)

# combine tourney seeds
df_tournament_seeds <- read_csv("data/NCAATourneySeeds.csv") %>% 
  inner_join(., df_teams, by = c("TeamID")) %>%
  select(-TeamID)
colnames(df_tournament_seeds) <- c("season", "seed", "team_name")


df_tournament_all <- df_tournament_all %>%
  inner_join(., df_tournament_seeds, by = c("season", "team_name"))

# add game rounds
df_tourney_seed_round_slots <- read_csv("data/NCAATourneySeedRoundSlots.csv")
colnames(df_tourney_seed_round_slots) <- c("seed", "game_round", "game_slot", "n_day_early", "n_day_late")

df_tournament_all <- df_tournament_all %>%
  inner_join(., df_tourney_seed_round_slots, by = c("seed"))
glimpse(df_tournament_all)

# filter and combine data
df_reg_tourney_round_1 <- df_tournament_all %>%
  ungroup() %>%
  filter(game_round == 5) %>%
  select(season, team_name, seed, outcome) %>%
  inner_join(., df_regular_season_all_total, by = c("season", "team_name")) %>%
  inner_join(., df_regular_season_all_mean, by = c("season", "team_name")) %>%
  inner_join(., df_reg_season_diff_total, by = c("season", "team_name")) %>%
  inner_join(., df_reg_season_diff_mean, by = c("season", "team_name")) %>%
  na.omit()
glimpse(df_reg_tourney_round_1)
```

### 2019 Regular Season
```{r}
df_regular_season_2019 <- read_csv("data/Prelim2019_RegularSeasonDetailedResults.csv") %>%
  inner_join(., df_wteams, by = "WTeamID") %>%
  inner_join(., df_lteams, by = "LTeamID") %>%
  select(-contains("TeamID")) %>%
  filter(Season == 2019)
glimpse(df_regular_season_2019)
```

### 2019 Seed
```{r}
seed_number_2019 <- readxl::read_excel("data/2019_NCAATourneySeeds.xlsx")
```

Let's break apart the tables into wins and losses so that we can manipulate the same variables at the same time.

```{r}
# separate wins and losses
df_regular_season_win_2019 <- df_regular_season_2019 %>%
  select(Season, DayNum, contains("W"))
colnames(df_regular_season_win_2019) <- gsub("W", "", colnames(df_regular_season_win_2019))

df_regular_season_loss_2019 <- df_regular_season_2019 %>%
  select(Season, DayNum, contains("L"))
colnames(df_regular_season_loss_2019) <- gsub("L", "", colnames(df_regular_season_loss_2019))

# include outcome
df_regular_season_win_2019 <- metric_data_manipulation(df_regular_season_win_2019) %>% 
  mutate(outcome = "win")

df_regular_season_loss_2019 <- metric_data_manipulation(df_regular_season_loss_2019) %>% 
  mutate(outcome = "loss")

df_win_num_2019 <- df_regular_season_win_2019 %>%
  select(-c(season, n_day, team_name, outcome, index))
    
df_loss_num_2019 <- df_regular_season_loss_2019 %>%
  select(-c(season, n_day, team_name, outcome, index))

df_reg_season_diff_2019 <- df_win_num_2019-df_loss_num_2019

df_reg_season_diff_2019 <- df_reg_season_diff_2019 %>%
  mutate(season = df_regular_season_win_2019$season,
         n_day = df_regular_season_win_2019$n_day,
         team_name = df_regular_season_win_2019$team_name,
         index = df_regular_season_win_2019$index)
glimpse(df_reg_season_diff_2019)

# combine together using bind_rows 
df_regular_season_all_2019 <- bind_rows(df_regular_season_win_2019, df_regular_season_loss_2019) %>% 
  mutate(outcome = as_factor(outcome))

glimpse(df_regular_season_all_2019)
```

Let's obtain overall season stats for each team. Let's obtain the total and mean.

```{r}
df_regular_season_all_total_2019 <- df_regular_season_all_2019 %>%
  select(-n_day, -index) %>%
  rowwise() %>%
  mutate(n_win = ifelse(outcome == "win", 1, 0)) %>%
  ungroup() %>%
  select(-outcome) %>%
  group_by(season, team_name) %>%
  summarise_all(sum) %>%
  ungroup()
glimpse(df_regular_season_all_total_2019)

df_regular_season_all_mean_2019 <- df_regular_season_all_2019 %>%
  select(-n_day, -index, -outcome) %>%
  group_by(season, team_name) %>%
  summarise_all(mean) %>%
  ungroup()
glimpse(df_regular_season_all_mean_2019)

df_reg_season_diff_total_2019 <- df_reg_season_diff_2019 %>%
  select(-c(index, n_day)) %>%
  group_by(season, team_name) %>%
  summarise_all(sum)
glimpse(df_reg_season_diff_total_2019)

df_reg_season_diff_mean_2019 <- df_reg_season_diff_2019 %>%
  select(-c(index, n_day)) %>%
  group_by(season, team_name) %>%
  summarise_all(mean)
glimpse(df_reg_season_diff_mean_2019)

colnames(df_regular_season_all_total_2019)[3:ncol(df_regular_season_all_total_2019)] <- paste0(colnames(df_regular_season_all_total_2019)[3:ncol(df_regular_season_all_total_2019)], "_total")

colnames(df_regular_season_all_mean_2019)[3:ncol(df_regular_season_all_mean_2019)] <- paste0(colnames(df_regular_season_all_mean_2019)[3:ncol(df_regular_season_all_mean_2019)], "_mean")

colnames(df_reg_season_diff_total_2019)[3:ncol(df_reg_season_diff_total_2019)] <- paste0(colnames(df_reg_season_diff_total_2019)[3:ncol(df_reg_season_diff_total_2019)], "_diff_total")

colnames(df_reg_season_diff_mean_2019)[3:ncol(df_reg_season_diff_mean_2019)] <- paste0(colnames(df_reg_season_diff_mean_2019)[3:ncol(df_reg_season_diff_mean_2019)], "_diff_mean")

df_reg_tourney_2019_all <- df_regular_season_all_total_2019 %>%
  inner_join(., df_regular_season_all_mean_2019, by = c("season", "team_name")) %>%
  inner_join(., df_reg_season_diff_total_2019, by = c("season", "team_name")) %>%
  inner_join(., df_reg_season_diff_mean_2019, by = c("season", "team_name"))
```

## LASSO Regression

### Set Outcome
```{r}
df_train_outcome <- df_reg_tourney_round_1 %>%
  select(outcome) %>%
  mutate(win = ifelse(outcome == "win", 1, 0)) %>%
  select(-outcome) %>%
  na.omit()
glimpse(df_train_outcome)
```

### Set Predictors
```{r}
df_train_predictors <- df_reg_tourney_round_1 %>%
  select(score_total:points_free_throws_made_percentage_diff_mean) %>%
  na.omit()
glimpse(df_train_predictors)
```

### Determine optimal lambda

Let's run a cross-validation of glm net to determine which lambda value we will need to use.
```{r}
cv_lasso <- cv.glmnet(x = as.matrix.data.frame(df_train_predictors), 
                      y = as.matrix.data.frame(df_train_outcome), 
                      alpha = 1, 
                      family = "binomial")
```

The lambda we should use is `r cv_lasso$lambda.min`.

### Model

Now, let's run the LASSO regression.

```{r}
model <- glmnet(as.matrix.data.frame(df_train_predictors), 
                as.matrix.data.frame(df_train_outcome), 
                alpha = 1, 
                family = "binomial",
                lambda = cv_lasso$lambda.min)
```

### Results

```{r}
coef(model)
```

## Predict Results

### Set Predictors

```{r}
df_test_predictors <- df_reg_tourney_2019_all %>%
  select(score_total:points_free_throws_made_percentage_diff_mean) %>%
  na.omit()
glimpse(df_test_predictors)
```

### Predict
```{r}
df_test_predicted <- model %>% 
  predict(newx = as.matrix.data.frame(df_test_predictors))
colnames(df_test_predicted) <- "predicted_outcome"

df_reg_tourney_2019_name <- df_reg_tourney_2019_all %>%
  na.omit() %>%
  select(team_name)

df_test_predicted_round_1 <- cbind(df_test_predicted, df_reg_tourney_2019_name) %>%
  inner_join(., seed_number_2019, by = "team_name") %>%
  arrange(n_seed)

df_test_predicted_round_1 %>% nice_tables()
```