---
title: "March Madness Analysis"
subtitle: "Tournament by Regular Season"
author: "Ekarin Eric Pongpipat, M.A."
date: "Date"
output:
  html_document:
    highlight: textmate
    theme: lumen
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

## R Packages
```{r}
packages <- c("tidyverse", "kableExtra", "broom", "broomExtra", "gganimate", "caret", "glmnet")
xfun::pkg_attach(packages, message = F)
```

## Functions
```{r}
# create nice html tables ----
nice_tables <- function(table) {
  table %>%
    kable() %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width = "100%", height = "200px")
}

# metric data manipulation ----
metric_data_manipulation <- function(data) {
  data %>%
  select(season = Season,
         n_day = DayNum,
         team_name = TeamName,
         score = Score,
         points_made = FGM,
         points_attempted = FGA,
         points_3_made = FGM3,
         points_3_attempted = FGA3,
         points_free_throws_made = FTM,
         points_free_throws_attempted = FTA,
         rebound_offensive = OR,
         rebound_defensive = DR,
         assist = Ast,
         turnovers = TO,
         steals = Stl,
         blocks = Blk,
         personal_fouls = PF) %>%
  mutate(index = 1:nrow(data),
         points_2_made = points_made - points_3_made,
         points_2_attempted = points_attempted - points_3_attempted,
         points_missed = points_attempted - points_made,
         points_2_missed = points_2_attempted - points_2_made,
         points_3_missed = points_3_attempted - points_3_made,
         points_free_throws_missed = points_free_throws_attempted - points_free_throws_made,
         points_made_percentage = points_made / points_attempted * 100,
         points_2_made_percentage = points_2_made / points_2_attempted * 100,
         points_3_made_percentage = points_3_made / points_3_attempted * 100,
         points_free_throws_made_percentage = points_free_throws_made / points_free_throws_attempted * 100)
}
```

## Regular Season
```{r}
df_regular_season <- read_csv("data/RegularSeasonDetailedResults.csv")

# add TeaamID and TeamName dataset
df_teams <- read_csv("data/Teams.csv")

# select only the TeamID and TeamName vars
df_teams <- df_teams %>% 
  select(TeamID, TeamName) 

# create df for win and loss TeamID and TeamNames to append to regular season df
df_wteams <- df_teams                         
colnames(df_wteams) <- c("WTeamID", "WTeamName")
df_lteams <- df_teams                        
colnames(df_lteams) <- c("LTeamID", "LTeamName")

# add actual names using inner_join
# remove TeamID
df_regular_season <- df_regular_season %>%
  inner_join(., df_wteams, by = "WTeamID") %>%
  inner_join(., df_lteams, by = "LTeamID") %>%
  select(-contains("TeamID"))

glimpse(df_regular_season)
```

Let's break apart the tables into wins and losses so that we can manipulate the same variables at the same time.

```{r}
# separate wins and losses
df_regular_season_win <- df_regular_season %>%
  select(Season, DayNum, contains("W"))
colnames(df_regular_season_win) <- gsub("W", "", colnames(df_regular_season_win))

df_regular_season_loss <- df_regular_season %>%
  select(Season, DayNum, contains("L"))
colnames(df_regular_season_loss) <- gsub("L", "", colnames(df_regular_season_loss))

# include outcome
df_regular_season_win <- metric_data_manipulation(df_regular_season_win) %>% 
  mutate(outcome = "win")

df_regular_season_loss <- metric_data_manipulation(df_regular_season_loss) %>% 
  mutate(outcome = "loss")

df_win_num <- df_regular_season_win %>%
  select(-c(season, n_day, team_name, outcome, index))
    
df_loss_num <- df_regular_season_loss %>%
  select(-c(season, n_day, team_name, outcome, index))

df_reg_season_diff <- df_win_num-df_loss_num

df_reg_season_diff <- df_reg_season_diff %>%
  mutate(season = df_regular_season_win$season,
         n_day = df_regular_season_win$n_day,
         team_name = df_regular_season_win$team_name,
         index = df_regular_season_win$index)
glimpse(df_reg_season_diff)

# combine together using bind_rows 
df_regular_season_all <- bind_rows(df_regular_season_win, df_regular_season_loss) %>% 
  mutate(outcome = as_factor(outcome))

glimpse(df_regular_season_all)
```

Let's obtain overall season stats for each team. Let's obtain the total and mean.

```{r}
df_regular_season_all_total <- df_regular_season_all %>%
  select(-n_day, -index) %>%
  rowwise() %>%
  mutate(n_win = ifelse(outcome == "win", 1, 0)) %>%
  ungroup() %>%
  select(-outcome) %>%
  group_by(season, team_name) %>%
  summarise_all(sum) %>%
  ungroup()
glimpse(df_regular_season_all_total)

df_regular_season_all_mean <- df_regular_season_all %>%
  select(-n_day, -index, -outcome) %>%
  group_by(season, team_name) %>%
  summarise_all(mean) %>%
  ungroup()
glimpse(df_regular_season_all_mean)

df_reg_season_diff_total <- df_reg_season_diff %>%
  select(-c(index, n_day)) %>%
  group_by(season, team_name) %>%
  summarise_all(sum)
glimpse(df_reg_season_diff_total)

df_reg_season_diff_mean <- df_reg_season_diff %>%
  select(-c(index, n_day)) %>%
  group_by(season, team_name) %>%
  summarise_all(mean)
glimpse(df_reg_season_diff_mean)
```

## Tourney

```{r}
df_tournament <- read_csv("data/NCAATourneyDetailedResults.csv") %>%
  inner_join(., df_wteams, by = "WTeamID") %>%
  inner_join(., df_lteams, by = "LTeamID") %>%
  select(-contains("TeamID"))
glimpse(df_tournament)

df_tournament_win <- df_tournament %>%
  select(Season, DayNum, contains("W"))
colnames(df_tournament_win) <- gsub("W", "", colnames(df_tournament_win))

df_tournament_loss <- df_tournament %>%
  select(Season, DayNum, contains("L"))
colnames(df_tournament_loss) <- gsub("L", "", colnames(df_tournament_loss))

df_tournament_win <- metric_data_manipulation(df_tournament_win) %>% 
  mutate(outcome = "win")
glimpse(df_tournament_win)

df_tournament_loss <- metric_data_manipulation(df_tournament_loss) %>% 
  mutate(outcome = "loss")
glimpse(df_tournament_loss)

df_tournament_all <- bind_rows(df_tournament_win, df_tournament_loss) %>% 
  mutate(outcome = as_factor(outcome))
glimpse(df_tournament_all)

df_tournament_all_total <- df_tournament_all %>%
  rowwise() %>%
  mutate(win = ifelse(outcome == "win", 1, 0)) %>%
  group_by(season, team_name) %>%
  summarise(n_win_tourney = sum(win)) %>%
  ungroup()
glimpse(df_tournament_all_total)

colnames(df_regular_season_all_total)[3:ncol(df_regular_season_all_total)] <- paste0(colnames(df_regular_season_all_total)[3:ncol(df_regular_season_all_total)], "_total")

colnames(df_regular_season_all_mean)[3:ncol(df_regular_season_all_mean)] <- paste0(colnames(df_regular_season_all_mean)[3:ncol(df_regular_season_all_mean)], "_mean")

colnames(df_reg_season_diff_total)[3:ncol(df_reg_season_diff_total)] <- paste0(colnames(df_reg_season_diff_total)[3:ncol(df_reg_season_diff_total)], "_diff_total")

colnames(df_reg_season_diff_mean)[3:ncol(df_reg_season_diff_mean)] <- paste0(colnames(df_reg_season_diff_mean)[3:ncol(df_reg_season_diff_mean)], "_diff_mean")

df_reg_tourney <- df_tournament_all_total %>%
  inner_join(., df_regular_season_all_total, by = c("season", "team_name")) %>%
  inner_join(., df_regular_season_all_mean, by = c("season", "team_name")) %>%
  inner_join(., df_reg_season_diff_total, by = c("season", "team_name")) %>%
  inner_join(., df_reg_season_diff_mean, by = c("season", "team_name"))
glimpse(df_reg_tourney)
```

```{r}
df_reg_tourney <- df_reg_tourney %>%
  na.omit()

df_train_predictors <- df_reg_tourney %>%
  filter(season != 2018) %>%
  select(score_total:points_free_throws_made_percentage_diff_mean) %>%
  na.omit()
glimpse(df_train_predictors)

df_train_outcome <- df_reg_tourney %>%
  filter(season != 2018) %>%
  select(n_win_tourney) %>%
  na.omit()
glimpse(df_train_outcome)
```

Let's run a cross-validation of glm net to determine which lambda value we will need to use.

```{r}
cv_lasso <- cv.glmnet(x = as.matrix.data.frame(df_train_predictors), 
                      y = as.matrix.data.frame(df_train_outcome), 
                      alpha = 1)
```

The lambda we should use is `r cv_lasso$lambda.min`.

Now, let's run the LASSO regression.

```{r}
model <- glmnet(as.matrix.data.frame(df_train_predictors), 
                as.matrix.data.frame(df_train_outcome), 
                alpha = 1,
                lambda = cv_lasso$lambda.min)
```

```{r}
coef(model)
```

```{r}
df_test_predictors <- df_reg_tourney %>%
  filter(season == 2018) %>%
  select(score_total:points_free_throws_made_percentage_diff_mean) %>%
  na.omit()
glimpse(df_test_predictors)

df_test_predicted <- model %>% 
  predict(newx = as.matrix.data.frame(df_test_predictors))
colnames(df_test_predicted) <- "predicted_outcome"

df_test_outcome <- df_reg_tourney %>%
  filter(season == 2018) %>%
  select(n_win_tourney) %>%
  na.omit()
colnames(df_test_outcome) <- "outcome"

df_validation <- cbind(df_test_outcome, df_test_predicted)

ggplot(df_validation, aes(outcome, predicted_outcome)) +
  geom_point()
```