---
title: "Untitled"
author: "Ekarin Eric Pongpipat, M.A."
date: "3/16/2019"
output: html_document
---

## R Libraries
```{r}
packages <- c("tidyverse", "caret", "glmnet")
xfun::pkg_attach(packages, message = F, install = T)
```

Let's first look at the regular season dataset.

## Regular Season
```{r}
df_regular_season <- read_csv("data/RegularSeasonDetailedResults.csv")
glimpse(df_regular_season)
```

We will also need to load the Teams.csv dataset to append actual team names rather than IDs.

```{r}
df_teams <- read_csv("data/Teams.csv")
glimpse(df_teams)

# select only the TeamID and TeamName vars
df_teams <- df_teams %>% select(TeamID, TeamName) 

# create df for win and loss TeamID and TeamNames to append to regular season df
df_wteams <- df_teams                         
colnames(df_wteams) <- c("WTeamID", "WTeamName")
df_lteams <- df_teams                        
colnames(df_lteams) <- c("LTeamID", "LTeamName")

# add actual names using inner_join
# remove TeamID
df_regular_season <- df_regular_season %>%
  inner_join(., df_wteams, by = "WTeamID") %>%
  inner_join(., df_lteams, by = "LTeamID") %>%
  select(-contains("TeamID"))

glimpse(df_regular_season)
```

Let's break apart the tables into wins and losses so that we can manipulate the same variables at the same time.

```{r}
df_regular_season_win <- df_regular_season %>%
  select(Season, DayNum, contains("W"))
colnames(df_regular_season_win) <- gsub("W", "", colnames(df_regular_season_win))

df_regular_season_loss <- df_regular_season %>%
  select(Season, DayNum, contains("L"))
colnames(df_regular_season_loss) <- gsub("L", "", colnames(df_regular_season_loss))

# data manipulation ----
# create new variables that are also easy to read
data_manipulation_reg_season <- function(data) {
  data %>%
  select(season = Season,
         n_day = DayNum,
         team_name = TeamName,
         score = Score,
         points_made = FGM,
         points_attempted = FGA,
         points_3_made = FGM3,
         points_3_attempted = FGA3,
         points_free_throws_made = FTM,
         points_free_throws_attempted = FTA,
         rebound_offensive = OR,
         rebound_defensive = DR,
         assist = Ast,
         turnovers = TO,
         steals = Stl,
         blocks = Blk,
         personal_fouls = PF) %>%
  mutate(index = 1:nrow(data),
         points_2_made = points_made - points_3_made,
         points_2_attempted = points_attempted - points_3_attempted,
         points_missed = points_attempted - points_made,
         points_2_missed = points_2_attempted - points_2_made,
         points_3_missed = points_3_attempted - points_3_made,
         points_free_throws_missed = points_free_throws_attempted - points_free_throws_made,
         points_made_percentage = points_made / points_attempted * 100,
         points_2_made_percentage = points_2_made / points_2_attempted * 100,
         points_3_made_percentage = points_3_made / points_3_attempted * 100,
         points_free_throws_made_percentage = points_free_throws_made / points_free_throws_attempted * 100)
}

df_regular_season_win <- data_manipulation_reg_season(df_regular_season_win) %>% mutate(outcome = "win")
glimpse(df_regular_season_win)

df_regular_season_loss <- data_manipulation_reg_season(df_regular_season_loss) %>% mutate(outcome = "loss")
glimpse(df_regular_season_loss)

df_regular_season_all <- bind_rows(df_regular_season_win, df_regular_season_loss) %>% 
  mutate(outcome = as_factor(outcome)) %>%
  na.omit()
glimpse(df_regular_season_all)
```

```{r}
df_design_matrix <- df_regular_season_all %>%
  select(score:points_free_throws_made_percentage, -index) %>%
  na.omit()
glimpse(df_design_matrix)

df_outcome <- df_regular_season_all %>%
  select(outcome) %>%
  mutate(win = ifelse(outcome == "win", 1, 0)) %>%
  select(win) %>%
  na.omit()
glimpse(df_outcome)
```

Let's run a cross-validation of glm net to determine which lambda value we will need to use.

```{r}
cv_lasso <- cv.glmnet(x = as.matrix.data.frame(df_design_matrix), 
                      y = as.matrix.data.frame(df_outcome), 
                      alpha = 1, 
                      family = "binomial")
```

The lambda we should use is `r cv_lasso$lambda.min`.

Now, let's run the LASSO regression.

```{r}
model <- glmnet(as.matrix.data.frame(df_design_matrix), 
                as.matrix.data.frame(df_outcome), 
                alpha = 1, 
                family = "binomial", 
                lambda = cv_lasso$lambda.min)
```

```{r}
coef(model)
```

## Tourney
```{r}
df_tournament <- read_csv("data/NCAATourneyDetailedResults.csv")
glimpse(df_tournament)
```

We will also need to load the Teams.csv dataset to append actual team names rather than IDs.

```{r}
df_teams <- read_csv("data/Teams.csv")
glimpse(df_teams)

# select only the TeamID and TeamName vars
df_teams <- df_teams %>% select(TeamID, TeamName) 

# create df for win and loss TeamID and TeamNames to append to regular season df
df_wteams <- df_teams                         
colnames(df_wteams) <- c("WTeamID", "WTeamName")
df_lteams <- df_teams                        
colnames(df_lteams) <- c("LTeamID", "LTeamName")

# add actual names using inner_join
# remove TeamID
df_tournament <- df_tournament %>%
  inner_join(., df_wteams, by = "WTeamID") %>%
  inner_join(., df_lteams, by = "LTeamID") %>%
  select(-contains("TeamID"))

glimpse(df_tournament)
```

Let's break apart the tables into wins and losses so that we can manipulate the same variables at the same time.

```{r}
df_tournament_win <- df_tournament %>%
  select(Season, DayNum, contains("W"))
colnames(df_tournament_win) <- gsub("W", "", colnames(df_tournament_win))

df_tournament_loss <- df_tournament %>%
  select(Season, DayNum, contains("L"))
colnames(df_tournament_loss) <- gsub("L", "", colnames(df_tournament_loss))

# data manipulation ----
# create new variables that are also easy to read
data_manipulation_tournament <- function(data) {
  data %>%
  select(season = Season,
         n_day = DayNum,
         team_name = TeamName,
         score = Score,
         points_made = FGM,
         points_attempted = FGA,
         points_3_made = FGM3,
         points_3_attempted = FGA3,
         points_free_throws_made = FTM,
         points_free_throws_attempted = FTA,
         rebound_offensive = OR,
         rebound_defensive = DR,
         assist = Ast,
         turnovers = TO,
         steals = Stl,
         blocks = Blk,
         personal_fouls = PF) %>%
  mutate(index = 1:nrow(data),
         points_2_made = points_made - points_3_made,
         points_2_attempted = points_attempted - points_3_attempted,
         points_missed = points_attempted - points_made,
         points_2_missed = points_2_attempted - points_2_made,
         points_3_missed = points_3_attempted - points_3_made,
         points_free_throws_missed = points_free_throws_attempted - points_free_throws_made,
         points_made_percentage = points_made / points_attempted * 100,
         points_2_made_percentage = points_2_made / points_2_attempted * 100,
         points_3_made_percentage = points_3_made / points_3_attempted * 100,
         points_free_throws_made_percentage = points_free_throws_made / points_free_throws_attempted * 100)
}

df_tournament_win <- data_manipulation_tournament(df_tournament_win) %>% mutate(outcome = "win")
glimpse(df_tournament_win)

df_tournament_loss <- data_manipulation_tournament(df_tournament_loss) %>% mutate(outcome = "loss")
glimpse(df_tournament_loss)

df_tournament_all <- bind_rows(df_tournament_win, df_tournament_loss) %>% mutate(outcome = as_factor(outcome))
glimpse(df_tournament_all)
```

```{r}
df_design_matrix_tourney <- df_tournament_all %>%
  select(score:points_free_throws_made_percentage, -index) %>%
  na.omit()
glimpse(df_design_matrix_tourney)

df_tourney_predicted_win_prob <- model %>% 
  predict(newx = as.matrix.data.frame(df_design_matrix_tourney))

df_tourney_predicted_outcome <- ifelse(df_tourney_win_prob > 0.5, "win", "loss")

df_tourney_outcome <- df_tournament_all %>%
  select(outcome) %>%
  na.omit() %>%
  mutate(predicted_outcome = df_tourney_predicted_outcome) %>%
  rowwise() %>%
  mutate(correct = ifelse(outcome == predicted_outcome, T, F),
         signal_detection = case_when(
    outcome == "win" && predicted_outcome == "win" ~ "hit",
    outcome == "win" && predicted_outcome == "loss" ~ "miss",
    outcome == "loss" && predicted_outcome == "loss" ~ "cr",
    outcome == "loss" && predicted_outcome == "win" ~ "fa"
  ))

glimpse(df_tourney_outcome)

df_summary_correct <- df_tourney_outcome %>%
  group_by(correct) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(total = sum(n)) %>%
  rowwise() %>%
  mutate(percent = n/total) %>%
  ungroup()

df_summary_sdt <- df_tourney_outcome %>%
  group_by(signal_detection) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(total = sum(n)) %>%
  rowwise() %>%
  mutate(percent = n/total) %>%
  ungroup()
```
